Mathematical Aspects of the Capital Market

### 2017-08-11 papers/FX/PaperBody.tex
Added an abstract to the FX paper.

\begin{abstract}
We consider an investor with preferences that accord with Generalized
Disappointment Aversion (GDA). Such an investor cares about downside
risk and we assume he recognizes the heavy tail feature of asset return
distributions. We argue that when a market is dominated by rational
investors of this kind, the return distributions of equities that are
actively traded in this market may have nearly equal tail-indices due
to monotonicity of the GDA preference with respect to the tail index.

We estimate tail indices and scale parameters of S\&P 500 stocks and
test the hypothesis that two given stock return series have the same
tail index.
% We also show, in contrast, the scale parameters of the return
%distributions may differ hugely from one another.
%On the other hand, whether or not all equities in a multivariate model
%have the same tail-index is a dividing issue for multivariate GARCH
%models proposed in the literature. Therefore, it is important to analyze
%data of real equity returns and see how close to each other the
%tail-indices actually are.
%In this work empirical results are also presented and they appear
%to support the conclusion that the tail-indices are very similar,
%with respect to the confidence bands of estimation.
\end{abstract}

@@

in the literature about how heavy these tails really are. For example, Barndorff-Nielsen and Shephard
\cite{barndorff:shephard:2001} and Eberlein \cite{eberlein:2001} favor ``semi-heavy'' tails

@@

\P(X_t>x)\sim c_+ \,x^{-\alpha_{\rm up}}\quad\mbox{and}\quad \P(X_t<-x)\sim c_-\,x^{-\alpha_{\rm low}}\,,\quad\xto\,,

@@

where $c_{\pm}$, $\alpha_{\rm up}$ and $\alpha_{\rm low}$ are positive constants.\footnote{Here and in what
  follows, $f(x)\sim g(x)$ for positive \fct s $f$ and $g$ means 
that $f(x)/g(x)\to 1$ as $\xto$.} See for example, Embrechts et

@@

\cite{resnick:2007}.  
In the extreme value literature it is common to replace the constants $c_\pm$ by suitable
{\em slowly varying} \fct s; cf. Embrechts et
al. \cite{embrechts:klueppelberg:mikosch:1997}, Chapter 3. In this
paper, for the  
sake of argument, we stick to the condition \eqref{eq:1}.

@@

There are some good theoretical reasons for the appearance of power-law tails in situations where certain moments 

@@

Fr\'echet \ds\ $\Phi_{\alpha_{\rm up}}(x)=\exp(-x^{-\alpha_{\rm up}})$ for $x>0$, i.e., scaled maxima of an iid \seq\ $(X_t)$ with

@@

generalized Pareto \ds\ which is the limit \ds\ of the excesses of $X_t$ above high thresholds, i.e., for a suitable positive scaling \fct\ $a(u)$,

@@

\P((X_t-u)/a(u)>x \mid X_t>x) \to (1+ x/\alpha_{\rm up})^{-\alpha_{\rm up}}\,,\qquad u\to\infty\,.

>>>>

\begin{abstract}
We consider an investor with preferences that accord with Generalized
Disappointment Aversion (GDA). Such an investor cares about downside
risk and we assume he recognizes the heavy tail feature of asset return
distributions. We argue that when a market is dominated by rational
investors of this kind, the return distributions of equities that are
actively traded in this market may have nearly equal tail-indices due
to monotonicity of the GDA preference with respect to the tail index.
We give conditions upon which the GDA preference is monotone and hence
suggests an equal tail index for all actively traded stocks.

We also estimate tail indices and scale parameters of S\&P 500 stocks and
test the hypothesis that two given stock return series have the same
tail index. The results vary across different sectors of the index.

% We also show, in contrast, the scale parameters of the return
%distributions may differ hugely from one another.
%On the other hand, whether or not all equities in a multivariate model
%have the same tail-index is a dividing issue for multivariate GARCH
%models proposed in the literature. Therefore, it is important to analyze
%data of real equity returns and see how close to each other the
%tail-indices actually are.
%In this work empirical results are also presented and they appear
%to support the conclusion that the tail-indices are very similar,
%with respect to the confidence bands of estimation.
\end{abstract}

@@

in the literature about how heavy these tails really are. For example,
Barndorff-Nielsen and Shephard \cite{barndorff:shephard:2001} and
Eberlein \cite{eberlein:2001} favor ``semi-heavy'' tails

@@

\P(X_t>x)\sim c_+ \,x^{-\alpha_{\rm up}}\quad\mbox{and}\quad
\P(X_t<-x)\sim c_-\,x^{-\alpha_{\rm low}}\,,\quad\xto\,,

@@

where $c_{\pm}$, $\alpha_{\rm up}$ and $\alpha_{\rm low}$ are positive
constants.\footnote{Here and in what follows, $f(x)\sim g(x)$ for
  positive \fct s $f$ and $g$ means that $f(x)/g(x)\to 1$ as $\xto$.}
See for example, Embrechts et

@@

\cite{resnick:2007}. In the extreme value literature it is common to
replace the constants $c_\pm$ by suitable {\em slowly varying} \fct s;
cf. Embrechts et al. \cite{embrechts:klueppelberg:mikosch:1997},
Chapter 3. In this paper, for the sake of argument, we stick to the condition \eqref{eq:1}.

@@

There are some good theoretical reasons for the appearance of
power-law tails in situations where certain moments 

@@

Fr\'echet \ds\ $\Phi_{\alpha_{\rm up}}(x)=\exp(-x^{-\alpha_{\rm up}})$
for $x>0$, i.e., scaled maxima of an iid \seq\ $(X_t)$ with

@@

generalized Pareto \ds\ which is the limit \ds\ of the excesses of
$X_t$ above high thresholds, i.e., for a suitable positive scaling
\fct\ $a(u)$,

@@

\P((X_t-u)/a(u)>x \mid X_t>x) \to (1+ x/\alpha_{\rm up})^{-\alpha_{\rm
    up}}\,,\qquad u\to\infty\,.

### 2017-08-11 phd/Chapterintroduction.tex
Added an abstract to the FX paper.

>>>>

\subsection{Importance sampling estimator of GARCH(p,q) rare event probability}
  We propose an efficient importance sampling estimator for the rare
  event probability $\P(|V| > u)$ where $V$ is a vector following the
  stationary distribution of a GARCH($p, q$) process. Recall
  $V_t = A_t V_{t-1} + B$ is the matrix recurrence equation of a
  GARCH($p, q$) process. We emanate the process $\{V_t\}$ from a set
  $\mathcal C = \{V: |V| \leq M\}$ for some predefined positive
  constant $M$, and then we introduce a dual change of measure for the
  originally iid matrices $\{A_t\}$: for
  $t < T_u = \min\{i > 0: |V_i| > u\}$, we exponentially tilt the
  distribution of $A_t | X_{t-1}$, where $X_0 = V_0$ and
  $X_t = \prod_{i=1}^t X_0/|\prod_{i=1}^t X_0|$, so that
  $|A_t X_{t-1}|^\xi$ is more likely to take on large values and hence
  $|V_t|$ is more likely to exceed $u$. Once the exeedance has
  happened, we change the
  distribution of $\{A_t\}$ back to the original and continue the
  process until $V_t$ returns to the set $\mathcal C$. Along each
  simulated path emenating from $\mathcal C$ and ending in
  $\mathcal C$ we compute $N_u = \sum_{i=1}^K \1{|V_i| > u}$, where
  $K = \min\{i > 0: |V_i| \leq M\}$.

  The pursuit estimate of $P(|V|>u)$ is then a weighted average of
  $N_u$ computed along each path. The weight depends on the path.

\subsection{Eigenvalues of the sample covariance matrix of a   
stochastic volatility model}
We consider a multivariate heavy-tailed stochastic volatility model
and analyze the large-sample behavior of its sample covariance
matrix. We study the limiting behavior of its entries in the
infinite-variance case and derive results for the ordered eigenvalues
and corresponding eigenvectors. Essentially, we consider two different
cases where the tail behavior either stems from the iid\ innovations
of the process or from its volatility sequence. In both cases, we make
use of a large deviations technique for regularly varying time series
to derive multivariate $\alpha$-stable limit distributions of the
sample covariance matrix. While we show that in the case of
heavy-tailed innovations the limiting behavior resembles that of
completely independent observations, we also derive that in the case
of a heavy-tailed volatility sequence the possible limiting behavior
is more diverse, i.e.\ allowing for dependencies in the limiting
distributions which are determined by the structure of the underlying
volatility sequence.

\subsection{Extreme value analysis for the sample autocovariance
  matrices of time series}
We  provide some  asymptotic theory for the largest eigenvalues of a
sample covariance matrix of a $p$-dimensional \ts\ where the dimension
$p=p_n$ converges to infinity when the sample size $n$ increases. 
We give a short overview of the literature on the topic both in the
light- and heavy-tailed cases when the data have finite (infinite)
fourth moment, respectively. Our main focus is on the heavy-tailed
case. In this case, one has a theory for the \pp\ of the normalized
eigenvalues of the sample covariance matrix in the iid case but also
when rows and columns of the data are linearly dependent. We provide
limit results for the weak \con\ of these \pp es to Poisson or cluster
Poisson processes. Based on this \con\ we can also derive the limit
laws of various \fct als of the ordered eigenvalues such as the 
joint \con\ of a finite number of the largest order statistics, the
joint limit law of the largest eigenvalue and the trace, limit laws
for successive ratios of ordered eigenvalues, etc. We also develop
some limit theory for the singular values of the sample autocovariance
matrices and their sums of squares. The theory is illustrated for
simulated data and for the components of the S\&P 500 stock index.
